###### QC, integration, annotation of heart samples
###### Original script made by Gino Bonazza, adapted by Maks Necki for microtissue data. 
######  17 Februrary, 2026

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 

### setup

```{r}
# Get current file name to make folder
current_file <- "QC_integration_annotation_heart"

# Load libraries
library(here)
library(readr)
library(readxl)
library(openxlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(speckle)
library(ggrepel)
library(edgeR)

#Output paths
output_dir_data <- here::here("scripts", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("scripts", "figure"))

output_dir_figs <- here::here("scripts", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)

```

## Quality control

```{r}
sample_list <- c(
  "Sample1/count/sample1_filtered_feature_bc_matrix.h5",
  "Sample2/count/sample2_filtered_feature_bc_matrix.h5",
  "Sample3/count/sample3_filtered_feature_bc_matrix.h5",
  "Sample4/count/sample4_filtered_feature_bc_matrix.h5"
)

base_dir <- "/mnt/dysk_D/NIKARD_35_probek/demultiplex_results/cDNA4_OCM_demux/outs/per_sample_outs/"

sample_names <- c("CZD5 muscle", "CZD16 muscle", "LVAD9 muscle", "LVAD73 muscle")

seurat_list <- lapply(seq_along(sample_list), function(i) {
  counts <- Read10X_h5(file.path(base_dir, sample_list[i]))
  obj <- CreateSeuratObject(counts = counts, project = "microtissue_analysis")
  obj$Sample <- sample_names[i]
  return(obj)
})

microtissue_preQC <- merge(seurat_list[[1]], y = seurat_list[-1],
                            add.cell.ids = sample_names,
                            project = "microtissue_analysis")
```

## Add metadata: percentage of mitochondrial and ribosomal genes

```{r}
microtissue_preQC[["percent.mt"]] <- PercentageFeatureSet(microtissue_preQC, pattern = "^MT-")
microtissue_preQC[["percent.rp"]] <- PercentageFeatureSet(microtissue_preQC, pattern = "^RP[SL]")
```


```{r}
microtissue_preQC
unique(microtissue_preQC$Sample) 
table(microtissue_preQC$Sample)
```

## Check quality control parameters

```{r}
p1 <- VlnPlot(microtissue_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(microtissue_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(microtissue_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(microtissue_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots

```



```{r}
p1 <- VlnPlot(microtissue_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(microtissue_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(microtissue_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(microtissue_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots_zoom

```


## Doublet detection


```{r}

DefaultAssay(microtissue_preQC)

Layers(microtissue_preQC[["RNA"]])

microtissue_preQC <- JoinLayers(microtissue_preQC)

Layers(microtissue_preQC[["RNA"]])

```




```{r}
# Konwersja Seurat -> SCE
microtissue_preQC_sce <- as.SingleCellExperiment(microtissue_preQC)

# Uruchomienie scDblFinder
microtissue_preQC_sce <- scDblFinder(
  microtissue_preQC_sce,
  samples = "Sample",   # kolumna w metadata z nazwami próbek
  clusters = FALSE       # jeśli masz jakieś klastry i chcesz je uwzględnić
)

# Sprawdzenie wyników
table(microtissue_preQC_sce@colData$scDblFinder.class)
```

```{r}
microtissue_preQC_sce <- logNormCounts(microtissue_preQC_sce)
```

```{r}
assayNames(microtissue_preQC_sce)
```


```{r}
microtissue_preQC <- as.Seurat(
  microtissue_preQC_sce,
  counts = "counts",    # surowe dane
  data = "logcounts"    # normalizowane log-transformed dane
)

rm(microtissue_preQC_sce)
```


```{r}
singlet_count <- table(microtissue_preQC$scDblFinder.class)["singlet"]
doublet_count <- table(microtissue_preQC$scDblFinder.class)["doublet"]

microtissue_preQC@meta.data[microtissue_preQC$scDblFinder.class == "singlet", "scDblFinder.n"] <- 
  paste0("Singlets (n=", singlet_count, ")")

microtissue_preQC@meta.data[microtissue_preQC$scDblFinder.class == "doublet", "scDblFinder.n"] <- 
  paste0("Doublets (n=", doublet_count, ")")

```


```{r}
levels_order <- c(
  paste0("Doublets (n=", doublet_count, ")"),
  paste0("Singlets (n=", singlet_count, ")")
)

microtissue_preQC$scDblFinder.n <- factor(
  microtissue_preQC$scDblFinder.n,
  levels = levels_order
)
```

```{r}
p1 <- VlnPlot(microtissue_preQC, features = c("nCount_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 50000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(microtissue_preQC, features = c("nFeature_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) 

QC_pre_Doublets <- p1 + p2 + plot_layout(ncol = 2)

QC_pre_Doublets
```

```{r}
output_dir_data <- here::here(current_file)

if (!dir.exists(output_dir_data)) {
  dir.create(output_dir_data, recursive = TRUE)
}
```


```{r}
saveRDS(microtissue_preQC, 
        here::here(output_dir_data, "heart_preQC.rds"))

```


```{r}
heart_preQC <- readRDS(here::here(output_dir_data, "heart_preQC.rds"))

```

## Remove Doublets

```{r}
heart <- subset(x = heart_preQC, subset = scDblFinder.class == "singlet")
table(heart@meta.data$scDblFinder.class)
```

```{r}
rm(heart_preQC)
heart@assays[["RAW"]] <- NULL
DefaultAssay(heart) <- "RNA"
```

## Check quality control parameters after removing doublets


```{r}
p1 <- VlnPlot(microtissue, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(microtissue, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(microtissue, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(microtissue, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots
```


```{r}
p1 <- VlnPlot(microtissue, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(microtissue, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(microtissue, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(microtissue, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots_zoom
```

## Filter based on number of counts, features and percentage of mitochondrial genes

```{r}
heart_filtered <- subset(heart, subset = 
                   nFeature_RNA > 500 &
                   percent.mt < 1.5 &
                   nCount_RNA > 1000 &
                   nCount_RNA < 100000)
table(heart_filtered$Sample)
```

```{r}
cell_type_cols <- c("#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A", "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600")  
sample_cols <- DiscretePalette_scCustomize(num_colors = 29, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
batch_cols <- DiscretePalette_scCustomize(num_colors = 9, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
cluster_cols <- DiscretePalette_scCustomize(num_colors = 15, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
```


```{r}
PalettePlot(cell_type_cols) + 
  PalettePlot(sample_cols) + 
  PalettePlot(batch_cols) + 
  PalettePlot(cluster_cols)
```

## Check quality control parameters after filtering


```{r}
p1 <- VlnPlot(heart_filtered, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(heart_filtered, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(heart_filtered, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(heart_filtered, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots

```


```{r}
p1 <- VlnPlot(heart_filtered, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(heart_filtered, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(heart_filtered, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(heart_filtered, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 1.5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots_zoom

```

## Integration and clustering

##### Normalization and scaling using SCTransform, regressing out percentage of mitochondrial genes

## Clustering without integration:

```{r}
heart_sct <- heart_filtered

DefaultAssay(heart_sct) <- "RNA"

heart_sct <- SCTransform(heart_sct) |>
  RunPCA() |>
  FindNeighbors() |>
  RunUMAP(dims = 1:50) |>
  FindClusters(resolution = seq(0.1, 0.8, by=0.1))
```


```{r}
clustree::clustree(heart_sct@meta.data[,grep("SCT_snn_res", colnames(heart_sct@meta.data))],
                   prefix = "SCT_snn_res.")

```

## Integrate the samples

```{r}
heart <- SCTransform(heart_filtered) |>
  RunPCA(npcs = 50, verbose = FALSE)


heart <- RunHarmony(heart, 
                     group.by.vars = "Sample",
                     dims.use = 1:50, 
                     plot_convergence=TRUE, 
                     .options = harmony_options(
                       max.iter.cluster = 100,
                       )
                     )

```


```{r}
colnames(heart@meta.data)

```


## Clustering after integration

```{r}
#DefaultAssay(heart) <- "SCT"
heart <- RunUMAP(heart, dims = 1:50, reduction = "harmony")
#heart@graphs <- list()
heart <- FindNeighbors(heart, dims = 1:50, reduction = "harmony", graph.name = "harmony_snn")
heart <- FindClusters(heart, graph.name = "harmony_snn", resolution = seq(0.01, 0.08, by=0.01))

```

```{r}
clustree::clustree(heart@meta.data[,grep("harmony_snn_res", colnames(heart@meta.data))],
                   prefix = "harmony_snn_res.")
```

```{r}
DimPlot(heart_sct, reduction = "umap", shuffle = T, group.by = c("SCT_snn_res.0.01", "Sample"), ncol = 2, label=T, pt.size = .1) + ggtitle("SCTransform without harmony") -> p1
DimPlot(heart, reduction = "umap", shuffle = T, group.by = c("harmony_snn_res.0.01", "Sample"), ncol = 2, label=T, pt.size = .1) + ggtitle("SCTransform with harmony") -> p2

png("umap_harmony_heart.png", width = 1000, height = 800)
p1 + p2
dev.off()

png("umap_heart_harmony_v2.png", width = 1000, height = 800)
p2
dev.off()

p1 + p2

```


```{r}
DimPlot(heart, reduction = "umap", shuffle = T, group.by = c("harmony_snn_res.0.01"), ncol = 1, label=T, pt.size = .1) + ggtitle("SCTransform with harmony") -> p3 

png("heart_harmony_snn_res001_clustered.png", width = 1000, height = 800)
p3
dev.off()

```


## Find the markers that characterize each cell population

```{r}
DefaultAssay(heart) <- "RNA"
if (length(Assays(heart)@data) == 0 || length(Assays(heart)@data[[1]]) == 0) {
  heart <- NormalizeData(heart)
  heart <- FindVariableFeatures(heart)
}

Idents(heart) <- heart$harmony_snn_res.0.1

Markers <- FindAllMarkers(heart, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 1)

write.csv(Markers, here::here(output_dir_data, "Heart_harmony_01.csv"))

Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "Heart_harmony_01_top10.csv"))

Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "Heart_harmony_01_top3.csv"))
```


```{r}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])
Heatmap <- DoHeatmap(subset(heart, cells = sample(colnames(heart), 8000)), label = FALSE, draw.line = F, features = Markers_top10$gene, assay = "SCT") +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))

png("heatmap_clustered_harmony_heart.png", width = 1000, height = 800)
Heatmap
dev.off()


Heatmap
```

```{r}
saveRDS(heart_sct, 
        here::here(output_dir_data, "heart_sct.rds"))

```


```{r}
saveRDS(heart, 
        here::here(output_dir_data, "heart_harmony.rds"))

```

```{r}
integrated <- readRDS(here::here(output_dir_data, "heart_harmony.rds"))
#unintegrated <- readRDS(here::here(output_dir_data, "heart_sct.rds"))
```


```{r}
png("markers_heart_harmony_featureplot_ part1.png", width = 1000, height = 800)
FeaturePlot_scCustom(integrated, features = c("TNNI3K", "RYR2", "FGF12", "MYH6",
                                              "DCN",  "COL1A1", "FN1", "ACTA2", 
                                              "COMP", "POSTN", "VWF", "PECAM1", 
                                              "CDH5" , "CD34", "TAGLN", "CNN1"), num_columns = 4)
dev.off()






                                           png("markers_heart_harmony_featureplot_ part2.png", width = 1000, height = 800)
FeaturePlot_scCustom(integrated, features =   c("MYH11", "MYLK", "LMOD1", "ACTG2",
                                              "PTPRC", "RGS6", "NRXN1", "MRC1", 
                                              "RGS5", "IL18R1", "PARP8", "SKAP1", 
                                              "TPD52", "TXNDC5", "EGFLAM", "POU2AF1"), num_columns = 4)
dev.off()

```


```{r}
png("top10markers_SCT_muscle_featureplot.png", width = 1000, height = 800)
FeaturePlot_scCustom(unintegrated, features = c("TAGLN", "CNN1", "MYH11", "MYLK", "LMOD1", "ACTG2"), num_columns = 3)
dev.off()
```

###### Annotate cell clusters:

```{r}
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("0"), "cell_type"] = "Fibroblasts I"
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("1"), "cell_type"] = "Fibroblasts II"
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("2"), "cell_type"] = "Cardiomyocytes"
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("3"), "cell_type"] = "Epithelial cells"
integrated$cell_type <- factor(integrated$cell_type, levels = c(
                                                                        "Fibroblasts I",
                                                                        "Fibroblasts II",
                                                                        "Cardiomyocytes",
                                                                        "Epithelial cells"))
Idents(integrated) <- integrated$cell_type

```


```{r}
integrated <- ScaleData(integrated)
mapal <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
mapal <- rev(mapal[1:256])
cluster_palette <- c("#F8766D","#24B700", "#00ACFC", "#D575FE")
Heatmap <- DoHeatmap(integrated, draw.line = F, features = Markers_top10$gene, group.colors = cluster_palette) +
  guides(colour=FALSE) +
  scale_fill_gradientn(colours = mapal) +
  theme(text = element_text(size = 15), axis.text.y = element_text(size = 8)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))

png("heatmap_clustered_harmony_annotated.png", width = 1000, height = 800)
Heatmap
dev.off()
```


```{r}
p <- DimPlot(integrated, group.by = "cell_type", reduction = "umap", label = F, cols = cluster_palette) + 
  NoLegend() + 
  theme(axis.text=element_text(size=14, face = "bold"), axis.title = element_text(size = 18, face = "bold")) + 
  theme(plot.title = element_blank())
LabelClusters(p, id = "cell_type", fontface = "bold", size = 5.5, repel = T)

```


```{r}
DotPlot(integrated, assay = "SCT", features = rev(Markers_top3$gene), dot.scale = 5, cluster.idents = FALSE) +
  RotatedAxis() +
  coord_flip() +
  theme(axis.title = element_blank(), axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 13), legend.text = element_text(size = 9), legend.title = element_text(size = 11), plot.margin = unit(c(0, 0, 0, 0.1), 
                           "inches"))

```



```{r}
Markers_top5 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC))
Idents(integrated) <- factor(integrated$cell_type, levels = rev(levels(integrated$cell_type)))
DotPlot(integrated, assay = "SCT", features = Markers_top5$gene, dot.scale = 5) +
  RotatedAxis() +
  theme(axis.title = element_blank(), axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 13), legend.text = element_text(size = 9), legend.title = element_text(size = 11), plot.margin = unit(c(0, 0, 0, 0.1), 
                           "inches"))

```



```{r DGE cluster 1 vs cluster 0}
Idents(integrated) <- integrated$cell_type
fibro <- subset(integrated, idents = c("Fibroblasts I", "Fibroblasts II"))

# Porównanie
Idents(fibro) <- fibro$cell_type

markers_fibro <- FindMarkers(
  object = fibro,
  ident.1 = "Fibroblasts I",
  ident.2 = "Fibroblasts II",
  assay = "SCT",
  logfc.threshold = 0,
  min.pct = 0
)

markers_fibro$gene <- rownames(markers_fibro)

```


```{r vulcanoplot cluster 1 vs cluster 0 single-cell markers}
top20 <- markers_fibro %>%
  arrange(desc(abs(avg_log2FC))) %>%
  slice(1:20)


ggplot(markers_fibro, aes(x = avg_log2FC, y = -log10(p_val))) +
  geom_point(alpha=0.5, color="grey") +
  coord_cartesian(ylim=c(0,250)) +
  geom_point(data = subset(markers_fibro, p_val_adj < 0.05 & abs(avg_log2FC) > 0.25), 
             aes(x = avg_log2FC, y = -log10(p_val)), color="red") +
  geom_hline(yintercept=-log10(0.05), linetype="dashed", color="blue") +
  geom_vline(xintercept=c(-0.25, 0.25), linetype="dashed", color="blue") +
  geom_text_repel(data = top20,
                  aes(label = gene),
                  size=3,
                  max.overlaps=20) +
  theme_classic() +
  labs(title="Volcano plot: Cluster 0 vs Cluster 1",
       x="log2 Fold Change", y="-log10(p-value)")

```


## Pseudobulk DGE analysis


```{r pseudocounts}

fibro <- subset(integrated, subset = cell_type %in% c("Fibroblasts I", "Fibroblasts II"))

pseudocounts <- AggregateExpression(
  fibro,
  assays = "RNA",
  group.by = c("Sample", "cell_type")
)$RNA


#colnames(pseudocounts)

```


```{r getting metadata for pseudobulk}
sample <- sub("_.*", "", colnames(pseudocounts))
cell_type <- sub(".*_", "", colnames(pseudocounts))

metadata <- data.frame(
  sample = sample,
  cell_type = cell_type
)

```


```{r}
dge <- DGEList(counts = pseudocounts)
dge$samples$group <- metadata$cell_type


# filtr próbek
keep_samples <- dge$samples$lib.size > 5e4
dge <- dge[, keep_samples]

# filtr genów
keep_genes <- filterByExpr(dge, group = dge$samples$group)
dge <- dge[keep_genes, , keep.lib.sizes = FALSE]
```


```{r DGE pseudobulk cluster 1 vs cluster 0}

dge <- calcNormFactors(dge)
dge$samples$group <- make.names(dge$samples$group)

design <- model.matrix(~0 + dge$samples$group)
colnames(design) <- levels(factor(dge$samples$group))
```


```{r}
dge <- estimateDisp(dge, design)

fit <- glmFit(dge, design)

contrast <- makeContrasts(
  FibroblastsI_vs_II = `Fibroblasts.I` - `Fibroblasts.II`,
  levels = design
)

lrt <- glmLRT(fit, contrast = contrast)


```


```{r}
res <- topTags(lrt, n = Inf)$table
res$FDR <- p.adjust(res$PValue, method = "fdr")

```


```{r volcano plot pseudobulk}
res$gene <- rownames(res)

res$significant <- res$FDR < 0.05 & abs(res$logFC) > 1

top20 <- res[order(res$FDR), ][1:20, ]

vulcan <- ggplot(res, aes(x = logFC, y = -log10(FDR))) +
  geom_point(aes(color = significant), alpha = 0.6) +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "red")) +
  geom_text_repel(
    data = top20,
    aes(label = gene),
    size = 3
  ) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  theme_classic() +
  theme(legend.position = "none")

png("fibroI_vs_fibroII_pseudobulk_vulcano.png", width = 1000, height = 800)
vulcan
dev.off()
```


```{r cm2d vs cm3d}
# ============================================
# 1. Wybór komórek: tylko Cardiomyocytes 2D i 3D
# ============================================
cm <- subset(
  integrated, 
  subset = cell_type == "Cardiomyocytes" & Sample %in% c("CM_2D", "CM_FB_3D")
)

# ============================================
# 2. Tworzymy pseudobulka
# ============================================
# Agregujemy ekspresję na poziomie próbek
pseudocounts <- AggregateExpression(
  cm,
  assays = "RNA",
  group.by = "Sample"
)$RNA

# ============================================
# 3. Tworzymy metadata do DGE
# ============================================
metadata <- data.frame(
  sample = colnames(pseudocounts),
  group  = colnames(pseudocounts)  # każda kolumna = grupa
)

dge <- DGEList(counts = pseudocounts)

metadata$group <- ifelse(grepl("CM-2D", metadata$sample), "CM-2D", "CM-FB-3D")
metadata$group <- factor(metadata$group, levels = c("CM-2D", "CM-FB-3D"))
dge$samples$group <- metadata$group
dge$samples$group <- make.names(dge$samples$group) 

# Filtr próbek i genów
keep_samples <- dge$samples$lib.size > 5e4
dge <- dge[, keep_samples]

keep_genes <- filterByExpr(dge, group = dge$samples$group)
dge <- dge[keep_genes, , keep.lib.sizes = FALSE]

# Normalizacja
dge <- calcNormFactors(dge)

```


```{r}
design <- model.matrix(~0 + dge$samples$group)
colnames(design) <- levels(dge$samples$group)

# Sprawdzenie, czy mamy dwa poziomy
levels(dge$samples$group)


contrast <- makeContrasts(CM2D_vs_CMFB3D = `CM.2D` - `CM.FB.3D`, levels = design)

# Estymacja i test
dge <- estimateDisp(dge, design)
fit <- glmFit(dge, design)
lrt <- glmLRT(fit, contrast = contrast)
res <- topTags(lrt, n=Inf)$table
res$FDR <- p.adjust(res$PValue, method="fdr")

```



```{r CM2D vs CM3D}
cm <- subset(integrated, subset = cell_type == "Cardiomyocytes" & Sample %in% c("CM_2D", "CM_FB_3D"))

pseudocounts <- AggregateExpression(
cm,
assays = "RNA",
group.by = c("Sample"),
)$RNA


metadata <- data.frame(
sample = colnames(pseudocounts),
group = colnames(pseudocounts) # nazwa grupy = nazwa sample
)

metadata$group <- factor(metadata$group, levels = c("CM-2D", "CM-3D"))

```


```{r}
dge <- DGEList(counts = pseudocounts)
dge$samples$group <- metadata$group


keep_samples <- dge$samples$lib.size > 5e4
dge <- dge[, keep_samples]


keep_genes <- filterByExpr(dge, group = dge$samples$group)
dge <- dge[keep_genes, , keep.lib.sizes = FALSE]

dge <- calcNormFactors(dge)

dge$samples$group <- make.names(dge$samples$group)
design <- model.matrix(~0 + dge$samples$group)
colnames(design) <- levels(factor(dge$samples$group))
```


```{r}
dge <- estimateDisp(dge, design)
fit <- glmFit(dge, design)

```

```{r}
contrast <- makeContrasts(
CM2D_vs_CM3D = CM.2D - CM.3D,
levels = design
)

lrt <- glmLRT(fit, contrast = contrast)


res <- topTags(lrt, n = Inf)$table
res$FDR <- p.adjust(res$PValue, method = "fdr")
res$gene <- rownames(res)

```