---
title: "Quick and easy pipeline for snRNAseq analysis of metformin induced microtissue"
subtitle: "Adapted from Gino Bonazza DCM samples analysis by Maks Nęcki and Magdalena Czepiec"
author:
 - Gino Bonazza
 - Maks Nęcki
 - Magdalena Czepiec
fontsize: 12 pt
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
---


# Introduction and importing files for analysis

## Global chunk visibility setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 


## Defining where to save analysis and interemediate files

```{r creating output folder}
current_file <- "QC_integration_annotation_metformin"

output_dir_data <- here::here(current_file)

if (!dir.exists(output_dir_data)) {
  dir.create(output_dir_data, recursive = TRUE)
}
```


### Importing necessary libraries

```{r libraries}
# Get current file name to make folder

# Load libraries
library(here)
library(readr)
library(readxl)
library(openxlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(ArchR)
library(clustree)
library(harmony)
library(speckle)
library(ggrepel)
library(edgeR)

#Output paths
output_dir_data <- here::here("scripts", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("scripts", "figure"))

output_dir_figs <- here::here("scripts", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)

```

## Importing cellranger and cellbender output files

```{r bender + ranger}
# -------------------------
# CELLBENDER
# -------------------------

cell_bender_merged <- Read_CellBender_h5_Multi_File(
  merge = TRUE,
  data_dir = "/mnt/dysk_D/NIKARD_35_probek/Lib11_OCM_demux/outs/per_sample_outs",
  custom_name = "/count/sample_clean_filtered.h5",
  sample_list = c(
    "Sample1",
    "Sample2",
    "Sample3",
    "Sample4"
  ),
  sample_names = c(
 "CM_NC", "CM_MET", "CM_TGF", "CM_MET_TGF"
  )
)

# -------------------------
# CELLRANGER
# -------------------------

cell_ranger_merged <- Read10X_h5_GEO(
  merge = TRUE,
  data_dir = "/mnt/dysk_D/NIKARD_35_probek/Lib11_OCM_demux/outs/per_sample_outs",
  shared_suffix = "/count/sample_filtered_feature_bc_matrix.h5",
  sample_list = c(
    "Sample1",
    "Sample2",
    "Sample3",
    "Sample4"
  ),
  sample_names = c(
    "CM_NC", "CM_MET", "CM_TGF", "CM_MET_TGF"
  )
)

```

## Merging cellranger with cellbender seurat objects into one big dataframe

```{r merge1} 

# -------------------------
# SEURAT OBJECT
# -------------------------

common_cells <- intersect(
  colnames(cell_ranger_merged),
  colnames(cell_bender_merged)
)

cell_ranger_merged <- cell_ranger_merged[, common_cells]
cell_bender_merged <- cell_bender_merged[, common_cells]

metformin_preQC  <- CreateSeuratObject(
  counts = cell_ranger_merged,
  assay = "RAW"
)

metformin_preQC[["RNA"]] <- CreateAssayObject(
  counts = cell_bender_merged
)

DefaultAssay(metformin_preQC) <- "RNA"
```


# Quality Control part 1 (mitochondrial and ribosomal genes marking + duplicates)


## Adding metadata: percentage of mitochondrial and ribosomal genes (by using regular expressions)

```{r mito + ribo}
metformin_preQC[["percent.mt"]] <- PercentageFeatureSet(metformin_preQC, pattern = "^MT-")
metformin_preQC[["percent.rp"]] <- PercentageFeatureSet(metformin_preQC, pattern = "^RP[SL]")
```

## Creating Sample column in our Seurat object (we substract from ident.orig string before first or second instance of "_" sign)

```{r sample}
metformin_preQC$Sample <- sub("_[^_]+$", "", colnames(metformin_preQC))

metformin_preQC
unique(metformin_preQC$Sample) 
table(metformin_preQC$Sample)
```

## First violinplot of number of cells, genes and percentage of mitochondrial and ribosomal RNA

```{r violin}
p1 <- VlnPlot(metformin_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots

```

## Zoom of the same violinplot (you can change parameters of zooming by modifying y.max parameter inside VLnPlot function)

```{r violin zoomed}
p1 <- VlnPlot(metformin_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 5) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 5) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots_zoom

```

## Doublet detection

```{r doublet}
# Conversion from Seurat format to SCE format (its is necessary for doublet detecion)
metformin_preQC_sce <- as.SingleCellExperiment(metformin_preQC)

# Running scDblFinder function (main function to detect doublets)
metformin_preQC_sce <- scDblFinder(
  metformin_preQC_sce,
  samples = "Sample",   
  clusters = FALSE     
)

# Checking the results - how many singlets and how many doublets
table(metformin_preQC_sce@colData$scDblFinder.class)
```

## Normalization (logarithm plus 1) of raw counts

```{r normalization}
metformin_preQC_sce <- logNormCounts(metformin_preQC_sce)
```


## Creation of new Seurat object: metformin_preQC, we add to this object info about doublets and once again we create "Sample" column

```{r merging with doublet information}
counts_mat <- assay(metformin_preQC_sce, "counts")
logcounts_mat <- assay(metformin_preQC_sce, "logcounts")

# we create new seurat object based on SCE dataframe
metformin_preQC <- CreateSeuratObject(counts = counts_mat, assay = "RNA")

# adding log-normalized layer
metformin_preQC <- SetAssayData(
  metformin_preQC,
  assay = "RNA",
  layer = "data",
  new.data = logcounts_mat
)

# switching to "RNA" as active assay/layer
DefaultAssay(metformin_preQC) <- "RNA"

# Adding doublet information
metformin_preQC@meta.data$scDblFinder.class <- colData(metformin_preQC_sce)$scDblFinder.class


# Once more adding sample information
metformin_preQC$Sample <- sub("_[^_]+$", "", colnames(metformin_preQC))
```

## factorization of doublet findings
```{r factorization}
metformin_preQC$scDblFinder.n <- factor(
  metformin_preQC$scDblFinder.n,
  levels = levels_order
)
```

## Checking how doublets and singlets are distributed across samples
```{r vulcano2}
p1 <- VlnPlot(metformin_preQC, features = c("nCount_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 50000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin_preQC, features = c("nFeature_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) 

QC_pre_Doublets <- p1 + p2 + plot_layout(ncol = 2)

QC_pre_Doublets
```

## First save. Checkpoint!
```{r checkpoint}
saveRDS(metformin_preQC, 
        here::here(output_dir_data, "metformin_ambient_preQC.rds"))

```

## We read saved .rds file

```{r read}
metformin_preQC <- readRDS(here::here(output_dir_data, "metformin_ambient_preQC.rds"))
```

## Lastly we remove doublets by `subset()` function

```{r subset}
metformin <- subset(x = metformin_preQC, subset = scDblFinder.class == "singlet")
table(metformin@meta.data$scDblFinder.class)
```

## Removing intermediate files and cleaning "RAW" layer

```{r}
rm(metformin_preQC)
metformin@assays[["RAW"]] <- NULL
DefaultAssay(metformin) <- "RNA"
```

## Checking quality control parameters after removing doublets

```{r}
metformin[["percent.mt"]] <- PercentageFeatureSet(metformin, pattern = "^MT-")
metformin[["percent.rp"]] <- PercentageFeatureSet(metformin, pattern = "^RP[SL]")


p1 <- VlnPlot(metformin, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots
```

## Filtering based on number of counts, features and percentage of mitochondrial genes

```{r subsetting}
metformin <- subset(metformin, subset = 
                   nFeature_RNA > 500 &
                   percent.mt < 0.5 & # For metformin we choose lower cutoff (you can change it to 1.5 for example)
                   nCount_RNA > 1000 & 
                   nCount_RNA < 100000)
table(metformin$Sample)
```

## Part from Gino code, we don't actually use it nor change it

```{r}
cell_type_cols <- c("#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A", "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600")  
sample_cols <- DiscretePalette_scCustomize(num_colors = 29, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
batch_cols <- DiscretePalette_scCustomize(num_colors = 9, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
cluster_cols <- DiscretePalette_scCustomize(num_colors = 15, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
```

## Same as above

```{r}
PalettePlot(cell_type_cols) + 
  PalettePlot(sample_cols) + 
  PalettePlot(batch_cols) + 
  PalettePlot(cluster_cols)
```

## Check quality control parameters after final filtering

```{r final vulcano}
p1 <- VlnPlot(metformin, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots

```


# Integration and clustering

## Integrate the samples (we use harmony function to integrate all four samples across library 11)

```{r integration}
metformin <- SCTransform(metformin) |>
  RunPCA(npcs = 50, verbose = FALSE)


metformin <- RunHarmony(metformin, # name of our dataset
                     group.by.vars = "Sample", #we want integrate by Sample
                     dims.use = 1:50, # dimensions from PCA, equal to npcs from RunPCA function. If you are certain that you have low variation across sample you can lower it to 1:20
                     plot_convergence=TRUE, # It draws figure how good samples are integrated
                     .options = harmony_options(
                       max.iter.cluster = 100, 
                       )
                     )

```


## Clustering after integration, we use kNN (k-nearest neiughbours) algorithm to find centroids of clusters, UMAP is short for uniform manifold approximation, it's just a fancy way to showcase PCA components

```{r clustering}
metformin <- RunUMAP(metformin, dims = 1:50, reduction = "harmony")
metformin <- FindNeighbors(metformin, dims = 1:50, reduction = "harmony", graph.name = "harmony_snn") # extremely important thing is to add graph.name hypervariable, without it your seurat object wouldnt have proper clustered columns
metformin <- FindClusters(metformin, graph.name = "harmony_snn", resolution = seq(0.1, 0.8, by=0.1))

```

## Visualizing how many clusters are produced from different resolutions (highest line is for smallest resolution and it is going step by step)
```{r}
clustree::clustree(metformin@meta.data[,grep("harmony_snn_res", colnames(metformin@meta.data))],
                   prefix = "harmony_snn_res.")
```

## Creation and saving of umap plot for sample and cluster resolution equal to 0.1 (in our case smallest)

```{r umap}

DimPlot(metformin, reduction = "umap", shuffle = T, group.by = c("harmony_snn_res.0.1", "Sample"), ncol = 2, label=T, pt.size = .1) + ggtitle("SCTransform with harmony") -> p2

png("umap_metformin_all_harmony_ambient_v2.png", width = 1000, height = 800)
p2
dev.off()

p2

```


## We see that cluster 3 is for tail genes and we decided upon analysing its features to delete it

```{r deleting}

metformin <- subset(metformin, subset = harmony_snn_res.0.1 != 3)


```

## OK, now let's merge cluster 0 and cluster 2. It was revealed from further analysis that they are the same cell type cluster (cardiomyocytes) so we decided to merge it

```{r merge}
metformin$harmony_snn_res.0.1[metformin$harmony_snn_res.0.1 == 2] <- 0
metformin$harmony_snn_res.0.1 <- droplevels(metformin$harmony_snn_res.0.1)
Idents(metformin) <- "harmony_snn_res.0.1"
```


## Find the markers that characterize each cell population. We are saving top10 top3 and all interesting biomarkes into .csv files

```{r biomarkers}

# If we don't at this stage have normalized data (e.g we don't have "data" slot in metadata) - we create it

DefaultAssay(metformin) <- "RNA"
if (length(Assays(metformin)@data) == 0 || length(Assays(metformin)@data[[1]]) == 0) {
  metformin <- NormalizeData(metformin)
  metformin <- FindVariableFeatures(metformin)
}

Idents(metformin) <- metformin$harmony_snn_res.0.1

Markers <- FindAllMarkers(metformin, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 1)

write.csv(Markers, here::here(output_dir_data, "Metformin__metformin_joined_ambient_harmony_v2_01.csv"))

Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "Metformin_metformin_joined_ambient_harmony_v2_01_top10.csv"))

Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "Metformin_metformin_joined_ambient_harmony_v2_01_top3.csv"))
```


## We create heatmap to show how good top10 biomarkers are adjusted across clusters


```{r heatmap}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])
Heatmap_metformin <- DoHeatmap(subset(metformin, cells = sample(colnames(metformin), 5000)), # Last number is variable. It depends on how much cells you have within your filtered top10 data file
label = FALSE, draw.line = F, features = Markers_top10$gene, assay = "SCT") +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))

png("heatmap_clustered_harmony_metformin_joined_ambient_metformin.png", width = 1000, height = 800)
Heatmap_metformin
dev.off()


Heatmap_metformin
```

## We save results of clustering and normalization. Checkpoint!

```{r save}
saveRDS(metformin, 
        here::here(output_dir_data, "metformin_cleaned.rds"))

```

## We read saved .rds file

```{r read2}
integrated <- readRDS(here::here(output_dir_data, "metformin_cleaned.rds"))
```

# Annotation (manual) and description of every cell type population within clusters


## We create `featureplot` too see expression of selected biomarkers in clusters. We can divide our clusters into two main populations: Cardiomyocytes and Fibroblasts
```{r}
png("markers_metformin_clustered_harmony_ambient_featureplot.png", width = 1000, height = 800)
FeaturePlot_scCustom(integrated, features = c("TNNI3K", "RYR2", "FGF12", "MYH6", #cardiomyocytes
                                              "DCN",  "COL1A1", "FN1", "ACTA2", # fibroblasts, myofibroblasts
                                              "COMP", "POSTN", "VWF", "PECAM1",  # endothelial cells
                                              "CDH5" , "CD34", "TAGLN", "CNN1", # immune-response fibroblasts
                                              "MYH11", "MYLK", "LMOD1", "ACTG2"), num_columns = 4) # various other biomarkers
dev.off()
```

## Annotate cell clusters:

```{r}
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("0"), "cell_type"] = "Fibromyocytes"
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("1"), "cell_type"] = "Fibroblasts"

integrated$cell_type <- factor(integrated$cell_type, levels = c("Fibromyocytes",
                                                                "Fibroblasts"))
Idents(integrated) <- integrated$cell_type

```

## We create heatmap but with text to which celltype which cluster belong

```{r heatmap2}
integrated <- ScaleData(integrated)
mapal <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
mapal <- rev(mapal[1:256])
cluster_palette <- c("#F8766D","#24B700")
Heatmap <- DoHeatmap(integrated, draw.line = F, features = Markers_top10$gene, group.colors = cluster_palette) +
  guides(colour=FALSE) +
  scale_fill_gradientn(colours = mapal) +
  theme(text = element_text(size = 15), axis.text.y = element_text(size = 8)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))

png("heatmap_clustered_harmony_annotated_metformin.png", width = 1000, height = 800)
Heatmap
dev.off()
```

## Same here but instead with UMAP, not heatmap:

```{r umap2}
p <- DimPlot(integrated, group.by = "cell_type", reduction = "umap", label = F, cols = cluster_palette) + 
  NoLegend() + 
  theme(axis.text=element_text(size=14, face = "bold"), axis.title = element_text(size = 18, face = "bold")) + 
  theme(plot.title = element_blank())
LabelClusters(p, id = "cell_type", fontface = "bold", size = 5.5, repel = T)

```

# Differential gene expression (using estimation of dispersion)

## First we create variable which will describe different conditions between our samples

```{r conditions}
integrated$condition <- integrated$Sample  
integrated$condition <- dplyr::recode(integrated$condition,
                                      "CM_NC"       = "NC",
                                      "CM_MET"      = "MET",
                                      "CM_TGF"      = "TGF",
                                      "CM_MET_TGF"  = "MET_TGF")

# Check
table(integrated$Sample, integrated$condition)
```

## Then we subset on which cell type we wnat first to check Differential Expression (DGE)

```{r DGE fibroblasts}
Idents(integrated) <- integrated$cell_type
Fibro <- subset(integrated, idents = c("Fibroblasts"))

# NC vs MET
Idents(Fibro) <- Fibro$condition
```

# Then we divide fibroblast gene data into two collections: First condition to test - healthy nc vs healthy cells with metformin induced, second condition to test: tgf induced ill cells with tgf cells with metformin

```{r}
# NC vs MET
deg_nc_met <- FindMarkers(
  Fibro,
  ident.1 = "NC",
  ident.2 = "MET",
  assay = "SCT",
  logfc.threshold = 0,
  min.pct = 0
)
deg_nc_met$gene <- rownames(deg_nc_met)

# TGF vs MET+TGF
deg_tgf <- FindMarkers(
  Fibro,
  ident.1 = "TGF",
  ident.2 = "MET_TGF",
  assay = "SCT",
  logfc.threshold = 0,
  min.pct = 0
)

deg_tgf$gene <- rownames(deg_tgf)
```

## Now we create contrast and design matrix and estimate dispersion. After estimation we use these three outputs to fit them to general linear model

```{r estimate disp}
# ----------------------------
# 1. Przygotowanie danych
# ----------------------------
# Załóżmy, że masz counts_matrix (raw counts) fibroblastów
# i metadata z kolumną 'condition' (np. NC, MET, TGF, MET_TGF)
counts <- as.matrix(GetAssayData(Fibro, assay = "SCT", layer = "counts"))

# wektor grup (musisz mieć dokładnie tyle elementów, ile kolumn w counts)
group <- factor(Fibro$condition)

# stwórz DGEList
dge <- DGEList(counts = counts, group = group)

# projekt modelu
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)

dge <- estimateDisp(dge, design)
fit <- glmFit(dge, design)

# kontrasty
contrast_nc_met <- makeContrasts(NC_vs_MET = NC - MET, levels = design)
contrast_tgf_met <- makeContrasts(TGF_vs_MET_TGF = TGF - `MET_TGF`, levels = design)

# LRT
lrt_nc_met <- glmLRT(fit, contrast = contrast_nc_met)
lrt_tgf_met <- glmLRT(fit, contrast = contrast_tgf_met)

```

## We convert `glmLRT` function output to simple dataframes for each experiment:

```{r simple data}
# ----------------------------
# 3. Wyniki do data.frame
# ----------------------------
deg_nc_met <- topTags(lrt_nc_met, n = Inf)$table %>%
  mutate(gene = rownames(.),
         fdr = p.adjust(PValue, method = "fdr"),
         avg_log2FC = logFC)

deg_tgf_met <- topTags(lrt_tgf_met, n = Inf)$table %>%
  mutate(gene = rownames(.),
         fdr = p.adjust(PValue, method = "fdr"),
         avg_log2FC = logFC)
```

## Now we define schema how we will show volcano plot function and genes on it

```{r volcano function}
# ----------------------------
# 4. Volcano plot
# ----------------------------
volcano_plot <- function(df, title, logfc_cut = 0.25, fdr_cut = 0.05, top_n = 20) {
  
  df <- df %>%
    mutate(Significant = ifelse(fdr < fdr_cut & abs(avg_log2FC) > logfc_cut,
                                ifelse(avg_log2FC > 0, "Up", "Down"), "ns"))
  
  top_genes <- df %>%
    filter(Significant != "ns") %>%
    arrange(desc(abs(avg_log2FC))) %>%
    slice_head(n = top_n)
  
  ggplot(df, aes(x = avg_log2FC, y = -log10(fdr))) +
    geom_point(aes(color = Significant), alpha = 0.5) +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "ns" = "grey")) +
    geom_hline(yintercept = -log10(fdr_cut), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-logfc_cut, logfc_cut), linetype = "dashed", color = "black") +
    geom_text_repel(data = top_genes, aes(label = gene), size = 3, max.overlaps = 20) +
    theme_classic() +
    labs(title = title, x = "log2 Fold Change", y = "-log10(FDR)")
}
```


## Lastly we run volcano function on our two dataframes from two conditions, we save it to png file.

```{r}
# Wyświetlenie
p1 <- volcano_plot(deg_nc_met, "NC vs MET")
p2 <- volcano_plot(deg_tgf_met, "TGF vs MET+TGF")

library(patchwork)
#p1 | 
p1 | p2

# Zapis do pliku
png("volcano_edgeR_metformin_Fibromyocytes.png", width = 1600, height = 800)
p1 | p2
dev.off()
```


# GSEA - gene set enrichment analysis

```{r}
gene_rank_df <- deg_tgf_met %>%
  dplyr::select(gene, avg_log2FC) %>%
  dplyr::rename(lfc = avg_log2FC) %>%
  dplyr::filter(!is.na(lfc)) %>%
  dplyr::distinct()

cat("Input genes:", nrow(gene_rank_df), "\n")


gene_rank <- gene_rank_df$lfc
names(gene_rank) <- gene_rank_df$gene

gene_rank <- sort(gene_rank, decreasing = TRUE)
```

```{r}
library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)


gene_map <- clusterProfiler::bitr(
  names(gene_rank),
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
)
```


```{r}
cat("Mapped genes:", nrow(gene_map), "\n")

```

```{r}
gene_rank_df <- gene_rank_df %>%
  inner_join(gene_map, by = c("gene" = "SYMBOL"))

# Create ranked vector
geneList <- gene_rank_df$lfc
names(geneList) <- gene_rank_df$ENTREZID

# remove duplicates FIRST (critical)
geneList <- geneList[!duplicated(names(geneList))]

# sort decreasing (required for GSEA)
geneList <- sort(geneList, decreasing = TRUE)

cat("Genes in ranked list:", length(geneList), "\n")

```

```{r}
# safety check
if (length(geneList) < 50) {
  warning("Very small ranked list — GSEA may be unstable")
}


# Run GSEA

GSEA_GO <- tryCatch(
  gseGO(
    geneList      = geneList,
    OrgDb         = org.Hs.eg.db,
    ont           = "BP",
    minGSSize     = 5,
    maxGSSize     = 1000,
    pvalueCutoff  = 0.05,
    verbose       = FALSE,
    eps           = 0,
    nPermSimple   = 20000
  ),
  error = function(e) {
    message("GSEA failed: ", e$message)
    NULL
  }
)


# Compute term similarity

if (!is.null(GSEA_GO) && nrow(as.data.frame(GSEA_GO)) > 0) {
  GSEA_GO <- pairwise_termsim(GSEA_GO)
  cat("GSEA terms:", nrow(GSEA_GO@result), "\n")
} else {
  message("No significant GSEA terms found.")
}

```


```{r}

activated_terms <- subset(GSEA_GO@result, NES > 0)
suppressed_terms <- subset(GSEA_GO@result, NES < 0)

# keep full object structure
GSEA_GO_activated <- GSEA_GO
GSEA_GO_activated@result <- activated_terms

GSEA_GO_suppressed <- GSEA_GO
GSEA_GO_suppressed@result <- suppressed_terms

# recompute similarity 
GSEA_GO_activated <- pairwise_termsim(GSEA_GO_activated)
GSEA_GO_suppressed <- pairwise_termsim(GSEA_GO_suppressed)

library(enrichplot)
library(ggplot2)
library(patchwork)

# -----------------------------
# Common theme
# -----------------------------
emap_theme <- theme(
  plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
  legend.position = "right"
)

# -----------------------------
# Activated terms
# -----------------------------
p <- emapplot(
  GSEA_GO_activated,
  showCategory = min(40, nrow(GSEA_GO_activated@result)),
  size_category = 1.0,
  min_edge = 0.3,
  layout = "kk"
)

# ⭐ shrink ONLY term labels
p$layers[[which(sapply(p$layers, function(x) "GeomTextRepel" %in% class(x$geom)) )]]$aes_params$size <- 3

x <- p +
  scale_color_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    name = "NES"
  ) +
  ggtitle("Activated Terms - metformin vs nc (healthy)") +
  emap_theme

png("gsea_Fibro_tgf_clean_activated.png", width = 1600, height = 800)
print(x)
dev.off()

```


```{r}
p_sup <- emapplot(
  GSEA_GO_suppressed,
  showCategory = min(40, nrow(GSEA_GO_suppressed@result)),
  size_category = 1.0,
  min_edge = 0.3,
  layout = "kk"
)

# ⭐ shrink ONLY term labels
p_sup$layers[[which(sapply(p_sup$layers, function(x) "GeomTextRepel" %in% class(x$geom)) )]]$aes_params$size <- 3

x <- p_sup +
  scale_color_gradient2(
    low = "#3B4CC0",
    mid = "white",
    high = "#B40426",
    midpoint = 0,
    name = "NES"
  ) +
  ggtitle("Suppressed Terms") +
  emap_theme

png("gsea_Fibro_tgf_clean_supressed.png", width = 1600, height = 800)
print(x)
dev.off()


```

