---
title: "Quick and easy pipeline for snRNAseq analysis of metformin induced microtissue"
subtitle: "Adapted from Gino Bonazza DCM samples analysis by Maks Nęcki and Magdalena Czepiec"
author:
 - Gino Bonazza
 - Maks Nęcki
 - Magdalena Czepiec
fontsize: 10 pt
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
---


# **Introduction and importing files for analysis**


## Global chunk visibility setup

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 


## Defining where to save analysis and interemediate files.

-.rmd and .csv files will be saved in *QC_integration_annotation_metformin* catalog \  if it's not created then this code will do it. 

```{r creating output folder}
current_file <- "QC_integration_annotation_metformin"

output_dir_data <- here::here(current_file)

if (!dir.exists(output_dir_data)) {
  dir.create(output_dir_data, recursive = TRUE)
}
```

## Importing necessary libraries

```{r}
# Function to load libraries with automatic installation
load_libraries <- function(lib_name, bioconductor = FALSE) {
  if (!require(lib_name, character.only = TRUE)) {
    if (bioconductor) {
      BiocManager::install(lib_name)
    } else {
      install.packages(lib_name)
    }
    require(lib_name, character.only = TRUE)
  }
}

# CRAN libraries
cran_libs <- c("here", "readr", "readxl", "openxlsx", "Matrix", "dplyr", 
               "ggplot2", "magrittr", "tidyverse", "reshape2", "pheatmap", 
               "png", "gridExtra", "knitr", "scales", "RColorBrewer", 
               "tibble", "patchwork", "statmod", "clustree", "harmony", 
               "ggrepel", "Matrix.utils")

# Bioconductor libraries
bioc_libs <- c("Seurat", "DropletUtils", "scDblFinder", "scCustomize", 
               "S4Vectors", "SingleCellExperiment", "scater", "ArchR", 
               "speckle", "edgeR")

# Load CRAN libraries
invisible(sapply(cran_libs, function(x) load_libraries(x, bioconductor = FALSE)))

# Load Bioconductor libraries
invisible(sapply(bioc_libs, function(x) load_libraries(x, bioconductor = TRUE)))
```

## Library descriptions:

| Name of library | Description/Function |
|:---------------:|:--------------------:|
| here | files check & save |
| readr | read & save *.rds* files |
| readxl | read & save *.csv* files |
| openxlsx | read & write Excel files |
| Seurat | single-cell RNA-seq analysis |
| DropletUtils | 10X data processing utilities |
| Matrix | sparse matrix operations |
| scDblFinder | doublet detection |
| scCustomize | scRNA-seq plot customization |
| dplyr | data manipulation |
| ggplot2 | data visualization |
| magrittr | pipe operator support |
| tidyverse | data science toolkit |
| reshape2 | data reshaping |
| S4Vectors | S4 vector structures |
| SingleCellExperiment | single-cell data container |
| pheatmap | heatmap visualization |
| png | PNG image handling |
| gridExtra | plot and table layouts |
| knitr | R Markdown rendering |
| scales | plot scaling utilities |
| RColorBrewer | color palettes |
| Matrix.utils | matrix data manipulation |
| tibble | modern data frames |
| scater | scRNA-seq QC & visualization |
| patchwork | plot composition |
| statmod | statistical modeling |
| ArchR | scATAC-seq analysis |
| clustree | cluster resolution visualization |
| harmony | batch correction & integration |
| speckle | cell proportion analysis |
| ggrepel | non-overlapping labels |
| edgeR | differential expression analysis |


# Importing and reading .RDS file after QC

```{r read}

metformin <- readRDS(
  "/media/maksn/Nowy1/demultiplex_results/scripts/metformin_final_QC.rds"
)


```


# Integration and clustering

## Integrate the samples (we use harmony function to integrate all four samples across library 11)

```{r integration}
metformin <- SCTransform(metformin) |>
  RunPCA(npcs = 50, verbose = FALSE)


metformin <- RunHarmony(metformin, # name of our dataset
                     group.by.vars = "Sample", #we want integrate by Sample
                     dims.use = 1:50, # dimensions from PCA, equal to npcs from RunPCA function. If you are certain that you have low variation across sample you can lower it to 1:20
                     plot_convergence=TRUE, # It draws figure how good samples are integrated
                     .options = harmony_options(
                       max.iter.cluster = 100, 
                       )
                     )

```


## Clustering after integration, we use kNN (k-nearest neiughbours) algorithm to find centroids of clusters, UMAP is short for uniform manifold approximation, it's just a fancy way to showcase PCA components

```{r clustering}
metformin <- RunUMAP(metformin, dims = 1:50, reduction = "harmony")
metformin <- FindNeighbors(metformin, dims = 1:50, reduction = "harmony", graph.name = "harmony_snn") # extremely important thing is to add graph.name hypervariable, without it your seurat object wouldnt have proper clustered columns
metformin <- FindClusters(metformin, graph.name = "harmony_snn", resolution = seq(0.1, 0.8, by=0.1))

```

## Visualizing how many clusters are produced from different resolutions (highest line is for smallest resolution and it is going step by step)
```{r}
clustree::clustree(metformin@meta.data[,grep("harmony_snn_res", colnames(metformin@meta.data))],
                   prefix = "harmony_snn_res.")
```

## Creation and saving of umap plot for sample and cluster resolution equal to 0.1 (in our case smallest)

```{r umap}

DimPlot(metformin, reduction = "umap", shuffle = T, group.by = c("harmony_snn_res.0.1", "Sample"), ncol = 2, label=T, pt.size = .1) + ggtitle("SCTransform with harmony") -> p2

png("umap_metformin_all_harmony_ambient_v2.png", width = 1000, height = 800)
p2
dev.off()

p2

```


```{r deleting}

metformin <- subset(metformin, subset = harmony_snn_res.0.1 != 3)
Idents(metformin) <- "harmony_snn_res.0.1"

```

```{r deleting}

metformin <- subset(metformin, subset = harmony_snn_res.0.1 != 2)
Idents(metformin) <- "harmony_snn_res.0.1"

```


## Let's now do quick sanity check whether the number of genes and cells across sample match everything:


```{r}

cat("Cells after integration: ", ncol(metformin), "\n\n")
cat("Genes after integration: ", nrow(metformin), "\n\n")


cat("\nCells per sample AFTER integration:\n")
print(table(metformin$Sample))

# opcjonalnie: sprawdzenie poziomów klastrów
cat("\nClusters AFTER integration:\n")
print(table(Idents(metformin)))


```


## Find the markers that characterize each cell population. We are saving top10 top3 and all interesting biomarkes into .csv files

```{r biomarkers}

# If we don't at this stage have normalized data (e.g we don't have "data" slot in metadata) - we create it

DefaultAssay(metformin) <- "RNA"
if (length(Assays(metformin)@data) == 0 || length(Assays(metformin)@data[[1]]) == 0) {
  metformin <- NormalizeData(metformin)
  metformin <- FindVariableFeatures(metformin)
}

Idents(metformin) <- metformin$harmony_snn_res.0.1

Markers <- FindAllMarkers(metformin, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 1) %>%
  arrange(cluster, desc(avg_log2FC))

head(Markers)

```


```{r}
write.csv(
  Markers,
  "/mnt/dysk_D/NIKARD_35_probek/demultiplex_results/scripts/QC_integration_annotation_metformin/Metformin_metformin_joined_ambient_harmony_v2_01.csv"
)

Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(
  Markers_top10,
  "/mnt/dysk_D/NIKARD_35_probek/demultiplex_results/scripts/QC_integration_annotation_metformin/Metformin_metformin_joined_ambient_harmony_v2_01_top10.csv"
)

Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(
  Markers_top3,
  "/mnt/dysk_D/NIKARD_35_probek/demultiplex_results/scripts/QC_integration_annotation_metformin/Metformin_metformin_joined_ambient_harmony_v2_01_top3.csv"
)
```



## We create heatmap to show how good top10 biomarkers are adjusted across clusters


```{r heatmap}
heatmap_cols <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
heatmap_cols <- rev(heatmap_cols[1:256])
Heatmap_metformin <- DoHeatmap(subset(metformin, cells = sample(colnames(metformin), 5000)), # Last number is variable. It depends on how much cells you have within your filtered top10 data file
label = FALSE, draw.line = F, features = Markers_top10$gene, assay = "SCT") +
  scale_fill_gradientn(colours = heatmap_cols) +
  theme(axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))

png("heatmap_clustered_harmony_metformin_joined_ambient_metformin.png", width = 1000, height = 800)
Heatmap_metformin
dev.off()


Heatmap_metformin
```

## We save results of clustering and normalization. Checkpoint!


```{r save}
saveRDS(
  metformin,
  "/media/maksn/Nowy1/demultiplex_results/scripts/metformin_final_clustered.rds"
)

```
