---
title: "Quick and easy pipeline for snRNAseq analysis of metformin induced microtissue"
subtitle: "Adapted from Gino Bonazza DCM samples analysis by Maks Nęcki and Magdalena Czepiec"
author:
 - Gino Bonazza
 - Maks Nęcki
 - Magdalena Czepiec
fontsize: 10 pt
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
---


# **Introduction and importing files for analysis**


## Global chunk visibility setup

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 


## Defining where to save analysis and interemediate files.

-.rmd and .csv files will be saved in *QC_integration_annotation_metformin* catalog \  if it's not created then this code will do it. 

```{r creating output folder}
current_file <- "QC_integration_metformin_clean"

output_dir_data <- here::here(current_file)

if (!dir.exists(output_dir_data)) {
  dir.create(output_dir_data, recursive = TRUE)
}
```


## Importing necessary libraries

Preferred version of this code would be with `require()` function since it checks\
if the library is installed and if not - it automatically installs and loads such libraries.\

```{r}
# Function to load libraries with automatic installation
load_libraries <- function(lib_name, bioconductor = FALSE) {
  if (!require(lib_name, character.only = TRUE)) {
    if (bioconductor) {
      BiocManager::install(lib_name)
    } else {
      install.packages(lib_name)
    }
    require(lib_name, character.only = TRUE)
  }
}

# CRAN libraries
cran_libs <- c("here", "readr", "readxl", "openxlsx", "Matrix", "dplyr", 
               "ggplot2", "magrittr", "tidyverse", "reshape2", "pheatmap", 
               "png", "gridExtra", "knitr", "scales", "RColorBrewer", 
               "tibble", "patchwork", "statmod", "clustree", "harmony", 
               "ggrepel", "Matrix.utils")

# Bioconductor libraries
bioc_libs <- c("Seurat", "DropletUtils", "scDblFinder", "scCustomize", 
               "S4Vectors", "SingleCellExperiment", "scater", "ArchR", 
               "speckle", "edgeR")

# Load CRAN libraries
invisible(sapply(cran_libs, function(x) load_libraries(x, bioconductor = FALSE)))

# Load Bioconductor libraries
invisible(sapply(bioc_libs, function(x) load_libraries(x, bioconductor = TRUE)))
```

## Library descriptions:

| Name of library | Description/Function |
|:---------------:|:--------------------:|
| here | files check & save |
| readr | read & save *.rds* files |
| readxl | read & save *.csv* files |
| openxlsx | read & write Excel files |
| Seurat | single-cell RNA-seq analysis |
| DropletUtils | 10X data processing utilities |
| Matrix | sparse matrix operations |
| scDblFinder | doublet detection |
| scCustomize | scRNA-seq plot customization |
| dplyr | data manipulation |
| ggplot2 | data visualization |
| magrittr | pipe operator support |
| tidyverse | data science toolkit |
| reshape2 | data reshaping |
| S4Vectors | S4 vector structures |
| SingleCellExperiment | single-cell data container |
| pheatmap | heatmap visualization |
| png | PNG image handling |
| gridExtra | plot and table layouts |
| knitr | R Markdown rendering |
| scales | plot scaling utilities |
| RColorBrewer | color palettes |
| Matrix.utils | matrix data manipulation |
| tibble | modern data frames |
| scater | scRNA-seq QC & visualization |
| patchwork | plot composition |
| statmod | statistical modeling |
| ArchR | scATAC-seq analysis |
| clustree | cluster resolution visualization |
| harmony | batch correction & integration |
| speckle | cell proportion analysis |
| ggrepel | non-overlapping labels |
| edgeR | differential expression analysis |

## Importing cellranger and cellbender output files

We have mapped results as .h5 file from `cellranger` command, we have also filtered in terms of ambient RNA removal results as a .h5 file from `cellbender`.\
We have four samples, Sample1, Sample2, Sample3 amd Sample4, they have appropriate names:\

| Sample | Condition |
|:------:|:---------:|
| Sample1 | CM_NC |
| Sample2 | CM_MET |
| Sample3 | CM_TGF |
| Sample4 | CM_MET_TGF |


```{r bender + ranger}

# -------------------------
# CELLBENDER
# -------------------------

cell_bender_merged <- Read_CellBender_h5_Multi_File(
  merge = TRUE,
  data_dir = "/mnt/dysk_D/NIKARD_35_probek/Lib11_OCM_demux/outs/per_sample_outs",
  custom_name = "/count/sample_clean_filtered.h5",
  sample_list = c(
    "Sample1",
    "Sample2",
    "Sample3",
    "Sample4"
  ),
  sample_names = c(
 "CM_NC", "CM_MET", "CM_TGF", "CM_MET_TGF"
  )
)

# -------------------------
# CELLRANGER
# -------------------------

cell_ranger_merged <- Read10X_h5_GEO(
  merge = TRUE,
  data_dir = "/mnt/dysk_D/NIKARD_35_probek/Lib11_OCM_demux/outs/per_sample_outs",
  shared_suffix = "/count/sample_filtered_feature_bc_matrix.h5",
  sample_list = c(
    "Sample1",
    "Sample2",
    "Sample3",
    "Sample4"
  ),
  sample_names = c(
    "CM_NC", "CM_MET", "CM_TGF", "CM_MET_TGF"
  )
)

```

## Merging cellranger with cellbender seurat objects into one big dataframe

In Seurat object we have set **RNA** as typical assay

```{r merge1} 

# -------------------------
# SEURAT OBJECT
# -------------------------

common_cells <- intersect(
  colnames(cell_ranger_merged),
  colnames(cell_bender_merged)
)

cell_ranger_merged <- cell_ranger_merged[, common_cells]
cell_bender_merged <- cell_bender_merged[, common_cells]

metformin_preQC  <- CreateSeuratObject(
  counts = cell_ranger_merged,
  assay = "RAW"
)

metformin_preQC[["RNA"]] <- CreateAssayObject(
  counts = cell_bender_merged
)

DefaultAssay(metformin_preQC) <- "RNA"


```


# Quality Control part 1 (mitochondrial and ribosomal genes marking + duplicates)


## Adding metadata: percentage of mitochondrial and ribosomal genes (by using regular expressions)

We add two columns into our dataframe: percentage of ribosomal and percentage of mitochondrial genes

```{r mito + ribo}
metformin_preQC[["percent.mt"]] <- PercentageFeatureSet(metformin_preQC, pattern = "^MT-")
metformin_preQC[["percent.rp"]] <- PercentageFeatureSet(metformin_preQC, pattern = "^RP[SL]")
```

## Creating Sample column in our Seurat object (we substract from ident.orig string before first or second instance of "_" sign)

### Check control - check number of genes and cells in each sample:

```{r sample}
metformin_preQC$Sample <- sub("_[^_]+$", "", colnames(metformin_preQC))

metformin_preQC
unique(metformin_preQC$Sample) 
table(metformin_preQC$Sample)
```

## First violinplot of number of cells, genes and percentage of mitochondrial and ribosomal RNA

```{r violin}
p1 <- VlnPlot(metformin_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots

```


## Doublet detection

```{r doublet, echo = FALSE}
# Conversion from Seurat format to SCE format (its is necessary for doublet detecion)
metformin_preQC_sce <- as.SingleCellExperiment(metformin_preQC)

# Running scDblFinder function (main function to detect doublets)
metformin_preQC_sce <- scDblFinder(
  metformin_preQC_sce,
  samples = "Sample",   
  clusters = FALSE     
)

# Checking the results - how many singlets and how many doublets
table(metformin_preQC_sce@colData$scDblFinder.class)
```

## Normalization (logarithm plus 1) of raw counts

```{r normalization}
metformin_preQC_sce <- logNormCounts(metformin_preQC_sce)
```


## Creation of new Seurat object: metformin_preQC, we add to this object info about doublets and once again we create "Sample" column

```{r merging with doublet information}
counts_mat <- assay(metformin_preQC_sce, "counts")
logcounts_mat <- assay(metformin_preQC_sce, "logcounts")

# we create new seurat object based on SCE dataframe
metformin_preQC <- CreateSeuratObject(counts = counts_mat, assay = "RNA")

# adding log-normalized layer
metformin_preQC <- SetAssayData(
  metformin_preQC,
  assay = "RNA",
  layer = "data",
  new.data = logcounts_mat
)

# switching to "RNA" as active assay/layer
DefaultAssay(metformin_preQC) <- "RNA"

# Adding doublet information
metformin_preQC@meta.data$scDblFinder.class <- colData(metformin_preQC_sce)$scDblFinder.class


# Once more adding sample information
metformin_preQC$Sample <- sub("_[^_]+$", "", colnames(metformin_preQC))
```

## Creating single/doublet flag column:

```{r flag1}
singlet_count <- table(metformin_preQC$scDblFinder.class)["singlet"]
doublet_count <- table(metformin_preQC$scDblFinder.class)["doublet"]

metformin_preQC@meta.data[metformin_preQC$scDblFinder.class == "singlet", "scDblFinder.n"] <- 
  paste0("Singlets (n=", singlet_count, ")")

metformin_preQC@meta.data[metformin_preQC$scDblFinder.class == "doublet", "scDblFinder.n"] <- 
  paste0("Doublets (n=", doublet_count, ")")

```


```{r flag2}
levels_order <- c(
  paste0("Doublets (n=", doublet_count, ")"),
  paste0("Singlets (n=", singlet_count, ")")
)

metformin_preQC$scDblFinder.n <- factor(
  metformin_preQC$scDblFinder.n,
  levels = levels_order
)
```

## factorization of doublet findings
```{r factorization}
metformin_preQC$scDblFinder.n <- factor(
  metformin_preQC$scDblFinder.n,
  levels = levels_order
)
```

## Checking how doublets and singlets are distributed across samples
```{r vulcano2}
p1 <- VlnPlot(metformin_preQC, features = c("nCount_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 50000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin_preQC, features = c("nFeature_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) 

QC_pre_Doublets <- p1 + p2 + plot_layout(ncol = 2)

QC_pre_Doublets
```

## First save. Checkpoint!
```{r checkpoint11}
saveRDS(
  metformin_preQC,
  "/mnt/dysk_D/NIKARD_35_probek/demultiplex_results/scripts/QC_integration_annotation_metformin/metformin_ambient_preQC.rds"
)

```

## We read saved .rds file

```{r read}
metformin_preQC <- readRDS(
  "/mnt/dysk_D/NIKARD_35_probek/demultiplex_results/scripts/QC_integration_annotation_metformin/metformin_ambient_preQC.rds"
)
```

## Lastly we remove doublets by `subset()` function

```{r subset}
metformin <- subset(x = metformin_preQC, subset = scDblFinder.class == "singlet")
table(metformin@meta.data$scDblFinder.class)
```

## Removing intermediate files and cleaning "RAW" layer

```{r}
#rm(metformin_preQC)
metformin@assays[["RAW"]] <- NULL
DefaultAssay(metformin) <- "RNA"
```

## Checking quality control parameters after removing doublets

```{r}
metformin[["percent.mt"]] <- PercentageFeatureSet(metformin, pattern = "^MT-")
metformin[["percent.rp"]] <- PercentageFeatureSet(metformin, pattern = "^RP[SL]")


p1 <- VlnPlot(metformin, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots
```

## Filtering based on number of counts, features and percentage of mitochondrial genes

```{r subsetting}
metformin <- subset(metformin, subset = 
                   nFeature_RNA > 200 &
                   percent.mt < 0.5 & # For metformin we choose lower cutoff (you can change it to 1.5 for example)
                   nCount_RNA > 1000 & 
                   nCount_RNA < 100000)
table(metformin$Sample)
```

## Part from Gino code, we don't actually use it nor change it, since we don't have batch or cluster cols for now

```{r}
cell_type_cols <- c("#d05050", "#33A02C", "#1F78B4", "#CAB2D6", "#FF7F00", "#B2DF8A", "#B15928", "#FFD000", "#FA95C0", "#FDBF6F", "#6A3D9A", "#7aCDF9", "#E6E600")  
sample_cols <- DiscretePalette_scCustomize(num_colors = 29, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
batch_cols <- DiscretePalette_scCustomize(num_colors = 9, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
cluster_cols <- DiscretePalette_scCustomize(num_colors = 15, palette = "polychrome", shuffle_pal = TRUE, seed = 220510)
```

## Same as above

```{r}
PalettePlot(cell_type_cols) + 
  PalettePlot(sample_cols) + 
  PalettePlot(batch_cols) + 
  PalettePlot(cluster_cols)
```

## Check quality control parameters after final filtering

```{r final vulcano}
p1 <- VlnPlot(metformin, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(metformin, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(metformin, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(metformin, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots

```

## Final sanity check:

```{r}


cat("Before QC:\n\n")
print(metformin_preQC)

table(metformin_preQC$Sample)

cat("After QC:\n\n")
print(metformin)

table(metformin$Sample)


```


## Final part of this code - saving. End of QC!
```{r checkpoint}
saveRDS(
  metformin,
  "/media/maksn/Nowy1/demultiplex_results/scripts/metformin_final_QC.rds"
)

```

