---
title: "Quick and easy pipeline for snRNAseq analysis of metformin induced microtissue"
subtitle: "Adapted from Gino Bonazza DCM samples analysis by Maks Nęcki and Magdalena Czepiec"
author:
 - Gino Bonazza
 - Maks Nęcki
 - Magdalena Czepiec
fontsize: 10 pt
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
---


# **Introduction and importing files for analysis**


## Global chunk visibility setup

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 


## Defining where to save analysis and interemediate files.

-.rmd and .csv files will be saved in *QC_integration_annotation_metformin* catalog \  if it's not created then this code will do it. 

```{r creating output folder}
current_file <- "QC_integration_annotation_metformin"

output_dir_data <- here::here(current_file)

if (!dir.exists(output_dir_data)) {
  dir.create(output_dir_data, recursive = TRUE)
}
```

## Importing necessary libraries

```{r}
# Function to load libraries with automatic installation
load_libraries <- function(lib_name, bioconductor = FALSE) {
  if (!require(lib_name, character.only = TRUE)) {
    if (bioconductor) {
      BiocManager::install(lib_name)
    } else {
      install.packages(lib_name)
    }
    require(lib_name, character.only = TRUE)
  }
}

# CRAN libraries
cran_libs <- c("here", "readr", "readxl", "openxlsx", "Matrix", "dplyr", 
               "ggplot2", "magrittr", "tidyverse", "reshape2", "pheatmap", 
               "png", "gridExtra", "knitr", "scales", "RColorBrewer", 
               "tibble", "patchwork", "statmod", "clustree", "harmony", 
               "ggrepel", "Matrix.utils")

# Bioconductor libraries
bioc_libs <- c("Seurat", "DropletUtils", "scDblFinder", "scCustomize", 
               "S4Vectors", "SingleCellExperiment", "scater", "ArchR", 
               "speckle", "edgeR")

# Load CRAN libraries
invisible(sapply(cran_libs, function(x) load_libraries(x, bioconductor = FALSE)))

# Load Bioconductor libraries
invisible(sapply(bioc_libs, function(x) load_libraries(x, bioconductor = TRUE)))
```

## Library descriptions:

| Name of library | Description/Function |
|:---------------:|:--------------------:|
| here | files check & save |
| readr | read & save *.rds* files |
| readxl | read & save *.csv* files |
| openxlsx | read & write Excel files |
| Seurat | single-cell RNA-seq analysis |
| DropletUtils | 10X data processing utilities |
| Matrix | sparse matrix operations |
| scDblFinder | doublet detection |
| scCustomize | scRNA-seq plot customization |
| dplyr | data manipulation |
| ggplot2 | data visualization |
| magrittr | pipe operator support |
| tidyverse | data science toolkit |
| reshape2 | data reshaping |
| S4Vectors | S4 vector structures |
| SingleCellExperiment | single-cell data container |
| pheatmap | heatmap visualization |
| png | PNG image handling |
| gridExtra | plot and table layouts |
| knitr | R Markdown rendering |
| scales | plot scaling utilities |
| RColorBrewer | color palettes |
| Matrix.utils | matrix data manipulation |
| tibble | modern data frames |
| scater | scRNA-seq QC & visualization |
| patchwork | plot composition |
| statmod | statistical modeling |
| ArchR | scATAC-seq analysis |
| clustree | cluster resolution visualization |
| harmony | batch correction & integration |
| speckle | cell proportion analysis |
| ggrepel | non-overlapping labels |
| edgeR | differential expression analysis |


# Importing and reading .rds file after integration/normalization and clusterization

```{r read}

integrated <- readRDS(
  "/media/maksn/Nowy1/demultiplex_results/scripts/metformin_final_annotated.rds"
)


```

# Differential gene expression (using estimation of dispersion) for fibroblasts

## First we create variable which will describe different conditions between our samples

```{r conditions}
integrated$condition <- integrated$Sample  
integrated$condition <- dplyr::recode(integrated$condition,
                                      "CM_NC"       = "NC",
                                      "CM_MET"      = "MET",
                                      "CM_TGF"      = "TGF",
                                      "CM_MET_TGF"  = "MET_TGF")

# Check
table(integrated$Sample, integrated$condition)
```

## Then we subset on which cell type we want first to check Differential Expression (DGE)

```{r}

Fibro <- subset(integrated, idents = "Fibroblasts")

Cardio <- subset(integrated, idents = "Cardiomyocytes")

```


```{r DGE fibroblasts}
Idents(Fibro) <- Fibro$condition
```

```{r DGE cardiomyocytes}
Idents(Cardio) <- Cardio$condition
```


# Then we divide fibroblast gene data into two collections: First condition to test - healthy nc vs healthy cells with metformin induced, second condition to test: tgf induced ill cells with tgf cells with metformin

```{r}

counts_cardio <- GetAssayData(Cardio, assay = "SCT", layer = "data")

# podział na grupy
cells_group1_cardio <- WhichCells(Cardio, idents = "NC")
cells_group2_cardio <- WhichCells(Cardio, idents = "MET")
cells_group3_cardio <- WhichCells(Cardio, idents = "TGF")
cells_group4_cardio <- WhichCells(Cardio, idents = "MET_TGF")

# procent ekspresji w każdej grupie
pct_group1_cardio <- rowSums(counts[, cells_group1] > 0) / length(cells_group1)
pct_group2_cardio <- rowSums(counts[, cells_group2] > 0) / length(cells_group2)
pct_group3_cardio <- rowSums(counts[, cells_group3] > 0) / length(cells_group3)
pct_group4_cardio <- rowSums(counts[, cells_group4] > 0) / length(cells_group4)

# filtr: ≥5% w jednej lub drugiej grupie

# Porównanie NC vs MET

genes_to_test_1_cardio <- names(pct_group1_cardio)[(pct_group1_cardio >= 0.1) | (pct_group2_cardio >= 0.1)]


# Porównanie TGF vs MET+TGF

genes_to_test_2_cardio <- names(pct_group2_cardio)[(pct_group2_cardio >= 0.1) | (pct_group4_cardio >= 0.1)]


# Porównanie NC vs TGF

genes_to_test_3_cardio <- names(pct_group1_cardio)[(pct_group1_cardio >= 0.1) | (pct_group3_cardio >= 0.1)]


```


```{r}
genes_to_plot_cardio <- c("VEGFA", "COL1A1", "TIMP3", "COL14A1", "VIM", "RIGI", "LRP1B", "SAT1")

VlnPlot(Cardio, features = genes_to_plot_cardio, group.by = "Sample", pt.size = 0.5, ncol = 2) +
  theme_minimal()

```


## Now we create contrast and design matrix and estimate dispersion. After estimation we use these three outputs to fit them to general linear model

```{r estimate disp}
# ----------------------------
# 1. Przygotowanie danych
# ----------------------------
# Załóżmy, że masz counts_matrix (raw counts) fibroblastów
# i metadata z kolumną 'condition' (np. NC, MET, TGF, MET_TGF)
counts_cardio <- as.matrix(GetAssayData(Cardio, assay = "RNA", layer = "counts"))



# wektor grup (musisz mieć dokładnie tyle elementów, ile kolumn w counts)
group_cardio <- factor(Cardio$condition)

# stwórz DGEList
dge_cardio <- DGEList(counts = counts_cardio, group = group_cardio)

# projekt modelu
design_cardio <- model.matrix(~0 + group_cardio)
colnames(design_cardio) <- levels(group_cardio)


contrast_met_nc_cardio <- makeContrasts(NC_vs_MET = MET - NC, levels = design_cardio) 
contrast_met_tgf_cardio <- makeContrasts(TGF_vs_MET_TGF = `MET_TGF` - TGF, levels = design_cardio) 
contrast_tgf_nc_cardio <- makeContrasts(NC_vs_TGF = TGF - NC, levels = design_cardio)


# MET vs NC


dge_met_nc_cardio <- dge[rownames(dge) %in% genes_to_test_1_cardio, , keep.lib.sizes = FALSE]

dge_met_nc_cardio <- estimateDisp(dge_met_nc_cardio, design_cardio)
fit_met_nc_cardio <- glmFit(dge_met_nc_cardio, design_cardio)

lrt_met_nc_cardio <- glmLRT(fit_met_nc_cardio, contrast = contrast_met_nc_cardio)


# MET_TGF vs TGF

dge_met_tgf_cardio <- dge_cardio[rownames(dge_cardio) %in% genes_to_test_2_cardio, , keep.lib.sizes = FALSE]

dge_met_tgf_cardio <- estimateDisp(dge_met_tgf_cardio, design_cardio)
fit_met_tgf_cardio <- glmFit(dge_met_tgf_cardio, design_cardio)

lrt_met_tgf_cardio <- glmLRT(fit_met_tgf_cardio, contrast = contrast_met_tgf_cardio)


# TGF vs NC

dge_tgf_nc_cardio <- dge_cardio[rownames(dge_cardio) %in% genes_to_test_3_cardio, , keep.lib.sizes = FALSE]

dge_tgf_nc_cardio <- estimateDisp(dge_tgf_nc_cardio, design_cardio)
fit_tgf_nc_cardio <- glmFit(dge_tgf_nc_cardio, design_cardio)

lrt_tgf_nc_cardio <- glmLRT(fit_tgf_nc_cardio, contrast = contrast_tgf_nc_cardio)


```

## We convert `glmLRT` function output to simple dataframes for each experiment:

```{r simple data}
deg_met_nc_cardio <- topTags(lrt_met_nc_cardio, n = Inf)$table %>%
  mutate(gene = rownames(.),
         avg_log2FC = logFC)

deg_met_tgf_cardio <- topTags(lrt_met_tgf_cardio, n = Inf)$table %>%
  mutate(gene = rownames(.),
         avg_log2FC = logFC)

deg_tgf_nc_cardio <- topTags(lrt_tgf_nc_cardio, n = Inf)$table %>%
  mutate(gene = rownames(.),
         avg_log2FC = logFC)
```

## Now we define schema how we will show volcano plot function and genes on it

```{r volcano function}
# ----------------------------
# 4. Volcano plot
# ----------------------------
volcano_plot <- function(df, title, logfc_cut = 0.25, fdr_cut = 0.05, top_n = 20) {
  
  df <- df %>%
    mutate(Significant = ifelse(FDR < fdr_cut & abs(avg_log2FC) > logfc_cut,
                                ifelse(avg_log2FC > 0, "Up", "Down"), "ns"))
  
  top_genes <- df %>%
    filter(Significant != "ns") %>%
    arrange(desc(abs(avg_log2FC))) %>%
    slice_head(n = top_n)
  
  ggplot(df, aes(x = avg_log2FC, y = -log10(FDR))) +
    geom_point(aes(color = Significant), alpha = 0.5) +
    scale_color_manual(values = c("Up" = "red", "Down" = "blue", "ns" = "grey")) +
    geom_hline(yintercept = -log10(fdr_cut), linetype = "dashed", color = "black") +
    geom_vline(xintercept = c(-logfc_cut, logfc_cut), linetype = "dashed", color = "black") +
    geom_text_repel(data = top_genes, aes(label = gene), size = 3, max.overlaps = 20) +
    theme_classic() +
    labs(title = title, x = "log2 Fold Change", y = "-log10(FDR)")
}
```


## Lastly we run volcano function on our two dataframes from two conditions, we save it to png file.

```{r}
# Wyświetlenie
p1_cardio <- volcano_plot(deg_met_nc_cardio, "NC vs MET")
p2_cardio <- volcano_plot(deg_met_tgf_cardio, "TGF vs MET+TGF")
p3_cardio <- volcano_plot(deg_tgf_nc_cardio, "NC vs TGF")

library(patchwork)
#p1 | 
p1_cardio | p2_cardio | p3_cardio

# Zapis do pliku
png("volcano_edgeR_metformin_Cardiomyocytes.png", width = 1600, height = 800)
p1_cardio | p2_cardio | p3_cardio
dev.off()
```





# Then we divide fibroblast gene data into two collections: First condition to test - healthy nc vs healthy cells with metformin induced, second condition to test: tgf induced ill cells with tgf cells with metformin

```{r}

counts_fibro <- GetAssayData(Fibro, assay = "SCT", layer = "data")

# podział na grupy
cells_group1_fibro <- WhichCells(Fibro, idents = "NC")
cells_group2_fibro <- WhichCells(Fibro, idents = "MET")
cells_group3_fibro <- WhichCells(Fibro, idents = "TGF")
cells_group4_fibro <- WhichCells(Fibro, idents = "MET_TGF")

# procent ekspresji w każdej grupie
pct_group1_fibro <- rowSums(counts_fibro[, cells_group1_fibro] > 0) / length(cells_group1_fibro)
pct_group2_fibro <- rowSums(counts_fibro[, cells_group2_fibro] > 0) / length(cells_group2_fibro)
pct_group3_fibro <- rowSums(counts_fibro[, cells_group3_fibro] > 0) / length(cells_group3_fibro)
pct_group4_fibro <- rowSums(counts_fibro[, cells_group4_fibro] > 0) / length(cells_group4_fibro)

# filtr: ≥5% w jednej lub drugiej grupie

# Porównanie NC vs MET

genes_to_test_1_fibro <- names(pct_group1_fibro)[(pct_group1_fibro >= 0.1) | (pct_group2_fibro >= 0.1)]


# Porównanie TGF vs MET+TGF

genes_to_test_2_fibro <- names(pct_group2_fibro)[(pct_group2_fibro >= 0.1) | (pct_group4_fibro >= 0.1)]


# Porównanie NC vs TGF

genes_to_test_3_fibro <- names(pct_group1_fibro)[(pct_group1_fibro >= 0.1) | (pct_group3_fibro >= 0.1)]


```


```{r}
genes_to_plot_fibro <- c("VEGFA", "COL1A1", "TIMP3", "COL14A1", "VIM", "RIGI", "LRP1B", "SAT1")

VlnPlot(Fibro, features = genes_to_plot_fibro, group.by = "Sample", pt.size = 0.5, ncol = 2) +
  theme_minimal()

```


## Now we create contrast and design matrix and estimate dispersion. After estimation we use these three outputs to fit them to general linear model

```{r estimate disp}
# ----------------------------
# 1. Przygotowanie danych
# ----------------------------
# Załóżmy, że masz counts_matrix (raw counts) fibroblastów
# i metadata z kolumną 'condition' (np. NC, MET, TGF, MET_TGF)
counts_fibro <- as.matrix(GetAssayData(Fibro, assay = "RNA", layer = "counts"))



# wektor grup (musisz mieć dokładnie tyle elementów, ile kolumn w counts)
group_fibro <- factor(Fibro$condition)

# stwórz DGEList
dge_fibro <- DGEList(counts = counts_fibro, group = group_fibro)

# projekt modelu
design_fibro <- model.matrix(~0 + group_fibro)
colnames(design_fibro) <- levels(group_fibro)


contrast_met_nc_fibro <- makeContrasts(NC_vs_MET = MET - NC, levels = design_fibro) 
contrast_met_tgf_fibro <- makeContrasts(TGF_vs_MET_TGF = `MET_TGF` - TGF, levels = design_fibro) 
contrast_tgf_nc_fibro <- makeContrasts(NC_vs_TGF = TGF - NC, levels = design_fibro)


# MET vs NC


dge_met_nc_fibro <- dge_fibro[rownames(dge_fibro) %in% genes_to_test_1_fibro, , keep.lib.sizes = FALSE]

dge_met_nc_fibro <- estimateDisp(dge_met_nc_fibro, design_fibro)
fit_met_nc_fibro <- glmFit(dge_met_nc_fibro, design_fibro)

lrt_met_nc_fibro <- glmLRT(fit_met_nc_fibro, contrast = contrast_met_nc_fibro)


# MET_TGF vs TGF

dge_met_tgf_fibro <- dge_fibro[rownames(dge_fibro) %in% genes_to_test_2_fibro, , keep.lib.sizes = FALSE]

dge_met_tgf_fibro <- estimateDisp(dge_met_tgf_fibro, design_fibro)
fit_met_tgf_fibro <- glmFit(dge_met_tgf_fibro, design_fibro)

lrt_met_tgf_fibro <- glmLRT(fit_met_tgf_fibro, contrast = contrast_met_tgf_fibro)


# TGF vs NC

dge_tgf_nc_fibro <- dge_fibro[rownames(dge_fibro) %in% genes_to_test_3_fibro, , keep.lib.sizes = FALSE]

dge_tgf_nc_fibro <- estimateDisp(dge_tgf_nc_fibro, design_fibro)
fit_tgf_nc_fibro <- glmFit(dge_tgf_nc_fibro, design_fibro)

lrt_tgf_nc_fibro <- glmLRT(fit_tgf_nc_fibro, contrast = contrast_tgf_nc_fibro)


```

## We convert `glmLRT` function output to simple dataframes for each experiment:

```{r simple data}
deg_met_nc_fibro <- topTags(lrt_met_nc_fibro, n = Inf)$table %>%
  mutate(gene = rownames(.),
         avg_log2FC = logFC)

deg_met_tgf_fibro <- topTags(lrt_met_tgf_fibro, n = Inf)$table %>%
  mutate(gene = rownames(.),
         avg_log2FC = logFC)

deg_tgf_nc_fibro <- topTags(lrt_tgf_nc_fibro, n = Inf)$table %>%
  mutate(gene = rownames(.),
         avg_log2FC = logFC)
```

## Now we define schema how we will show volcano plot function and genes on it


## Lastly we run volcano function on our two dataframes from two conditions, we save it to png file.

```{r}
# Wyświetlenie
p1_fibro <- volcano_plot(deg_met_nc_fibro, "NC vs MET")
p2_fibro <- volcano_plot(deg_met_tgf_fibro, "TGF vs MET+TGF")
p3_fibro <- volcano_plot(deg_tgf_nc_fibro, "NC vs TGF")

library(patchwork)
#p1 | 
p1_fibro | p2_fibro | p3_fibro

# Zapis do pliku
png("volcano_edgeR_metformin_Fibroblasts.png", width = 1600, height = 800)
p1_fibro | p2_fibro | p3_fibro
dev.off()

```


# Saving



## Saving to final csv-s cardiomyocytes results:

```{r}

write.csv(deg_met_nc_cardio, file = "deg_MET_vs_NC_cardio.csv", row.names = FALSE)
write.csv(deg_met_tgf_cardio, file = "deg_MET_TGF_vs_TGF_cardio.csv", row.names = FALSE)
write.csv(deg_tgf_nc_cardio, file = "deg_TGF_vs_NC_cardio.csv", row.names = FALSE)

```


## Saving to final csv-s fibroblasts results:

```{r}

write.csv(deg_met_nc_fibro, file = "deg_MET_vs_NC_fibro.csv", row.names = FALSE)
write.csv(deg_met_tgf_fibro, file = "deg_MET_TGF_vs_TGF_fibro.csv", row.names = FALSE)
write.csv(deg_tgf_nc_fibro, file = "deg_TGF_vs_NC_fibro.csv", row.names = FALSE)

```

