---
title: "Quick and easy pipeline for snRNAseq analysis of metformin induced microtissue"
subtitle: "Adapted from Gino Bonazza DCM samples analysis by Maks Nęcki and Magdalena Czepiec"
author:
 - Gino Bonazza
 - Maks Nęcki
 - Magdalena Czepiec
fontsize: 10 pt
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
---


# **Introduction and importing files for analysis**


## Global chunk visibility setup

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
``` 


## Defining where to save analysis and interemediate files.

-.rmd and .csv files will be saved in *QC_integration_annotation_metformin* catalog \  if it's not created then this code will do it. 

```{r creating output folder}
current_file <- "QC_integration_annotation_metformin"

output_dir_data <- here::here(current_file)

if (!dir.exists(output_dir_data)) {
  dir.create(output_dir_data, recursive = TRUE)
}
```

## Importing necessary libraries

```{r}
# Function to load libraries with automatic installation
load_libraries <- function(lib_name, bioconductor = FALSE) {
  if (!require(lib_name, character.only = TRUE)) {
    if (bioconductor) {
      BiocManager::install(lib_name)
    } else {
      install.packages(lib_name)
    }
    require(lib_name, character.only = TRUE)
  }
}

# CRAN libraries
cran_libs <- c("here", "readr", "readxl", "openxlsx", "Matrix", "dplyr", 
               "ggplot2", "magrittr", "tidyverse", "reshape2", "pheatmap", 
               "png", "gridExtra", "knitr", "scales", "RColorBrewer", 
               "tibble", "patchwork", "statmod", "clustree", "harmony", 
               "ggrepel", "Matrix.utils")

# Bioconductor libraries
bioc_libs <- c("Seurat", "DropletUtils", "scDblFinder", "scCustomize", 
               "S4Vectors", "SingleCellExperiment", "scater", "ArchR", 
               "speckle", "edgeR")

# Load CRAN libraries
invisible(sapply(cran_libs, function(x) load_libraries(x, bioconductor = FALSE)))

# Load Bioconductor libraries
invisible(sapply(bioc_libs, function(x) load_libraries(x, bioconductor = TRUE)))
```

## Library descriptions:

| Name of library | Description/Function |
|:---------------:|:--------------------:|
| here | files check & save |
| readr | read & save *.rds* files |
| readxl | read & save *.csv* files |
| openxlsx | read & write Excel files |
| Seurat | single-cell RNA-seq analysis |
| DropletUtils | 10X data processing utilities |
| Matrix | sparse matrix operations |
| scDblFinder | doublet detection |
| scCustomize | scRNA-seq plot customization |
| dplyr | data manipulation |
| ggplot2 | data visualization |
| magrittr | pipe operator support |
| tidyverse | data science toolkit |
| reshape2 | data reshaping |
| S4Vectors | S4 vector structures |
| SingleCellExperiment | single-cell data container |
| pheatmap | heatmap visualization |
| png | PNG image handling |
| gridExtra | plot and table layouts |
| knitr | R Markdown rendering |
| scales | plot scaling utilities |
| RColorBrewer | color palettes |
| Matrix.utils | matrix data manipulation |
| tibble | modern data frames |
| scater | scRNA-seq QC & visualization |
| patchwork | plot composition |
| statmod | statistical modeling |
| ArchR | scATAC-seq analysis |
| clustree | cluster resolution visualization |
| harmony | batch correction & integration |
| speckle | cell proportion analysis |
| ggrepel | non-overlapping labels |
| edgeR | differential expression analysis |

# Importing and reading .rds file after integration/normalization and clusterization

```{r read}

integrated <- readRDS(
  "/media/maksn/Nowy1/demultiplex_results/scripts/metformin_final_clustered.rds"
)


```

## We read saved .rds file


# Annotation (manual) and description of every cell type population within clusters


```{r}

integrated
genes <- rownames(integrated@assays$RNA)
length(genes)  # wszystkie geny w RNA assay


table(integrated$Sample)
```



## We create `featureplot` too see expression of selected biomarkers in clusters. We can divide our clusters into two main populations: Cardiomyocytes and Fibroblasts
```{r}
#png("markers_metformin_clustered_harmony_ambient_featureplot.png", width = 1000, height = 800)
FeaturePlot_scCustom(integrated, features = c("TNNI3K", "RYR2", "FGF12", "MYH6", #cardiomyocytes
                                              "DCN",  "COL1A1", "FN1", "ACTA2", # fibroblasts, myofibroblasts
                                              "COMP", "POSTN", "VWF", "PECAM1",  # endothelial cells
                                              "CDH5" , "CD34", "TAGLN", "CNN1", # immune-response fibroblasts
                                              "MYH11", "MYLK", "LMOD1", "ACTG2"), num_columns = 4) # various other biomarkers

```



## Annotate cell clusters:

```{r}
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("0"), "cell_type"] = "Cardiomyocytes"
integrated@meta.data[integrated@meta.data$harmony_snn_res.0.1 %in% c("1"), "cell_type"] = "Fibroblasts"

integrated$cell_type <- factor(integrated$cell_type, levels = c("Cardiomyocytes",
                                                                "Fibroblasts"))
Idents(integrated) <- integrated$cell_type

```

## We check the percentage of interesting biomarkers in each of sample in fibroblasts:

```{r subsetting fibro}


Fibro <- subset(integrated, idents = "Fibroblasts")



Cardio <- subset(integrated, idents = "Cardiomyocytes")

```


```{r}
genes_to_plot <- c("COL1A1", "POSTN", "COMP", "COL3A1", "ACTA2", "FN1", "TNC", "TIMP3")

VlnPlot(Cardio, features = genes_to_plot, group.by = "Sample", pt.size = 0.5, ncol = 2) +
  theme_minimal()

```

```{r}

VlnPlot(Cardio, features = genes_to_plot, group.by = "Sample", pt.size = 0.5, ncol = 2) +
  theme_minimal()

```

## We create heatmap but with text to which celltype which cluster belong

```{r heatmap2}
integrated <- ScaleData(integrated)
mapal <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
mapal <- rev(mapal[1:256])
cluster_palette <- c("#F8766D","#24B700")
Heatmapa <- DoHeatmap(integrated, draw.line = F, features = Markers_top10$gene, group.colors = cluster_palette) +
  guides(colour=FALSE) +
  scale_fill_gradientn(colours = mapal) +
  theme(text = element_text(size = 15), axis.text.y = element_text(size = 8)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))


#png("heatmap_clustered_harmony_annotated_metformin.png", width = 1000, height = 800)
Heatmapa
#dev.off()
```

## Same here but instead with UMAP, not heatmap:

```{r umap2}
p <- DimPlot(integrated, group.by = "cell_type", reduction = "umap", label = F, cols = cluster_palette) + 
  NoLegend() + 
  theme(axis.text=element_text(size=14, face = "bold"), axis.title = element_text(size = 18, face = "bold")) + 
  theme(plot.title = element_blank())
LabelClusters(p, id = "cell_type", fontface = "bold", size = 5.5, repel = T)

```


## Let's now do quick sanity check whether the number of genes and cells across sample match everything:


```{r}

cat("Cells after annotation: ", ncol(integrated), "\n\n")
cat("Genes after annotation: ", nrow(integrated), "\n\n")


cat("\nCells per sample AFTER annotation:\n")
print(table(integrated$Sample))

# opcjonalnie: sprawdzenie poziomów klastrów
cat("\nClusters AFTER annotation:\n")
print(table(Idents(integrated)))


```



```{r save}
saveRDS(
  integrated,
  "/media/maksn/Nowy1/demultiplex_results/scripts/metformin_final_annotated.rds"
)

```
